\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-10-03 paper/main.tex
CH: Editing the model section, adding some simple analytical observations to the results section

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress,round
]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

@@

\newcommand{\splitatcommas}[1]{%
  \begingroup
  \begingroup\lccode`~=`, \lowercase{\endgroup \edef~{\mathchar\the\mathcode`,
    \penalty0 \noexpand\hspace{0pt plus 1em}}%
  }\mathcode`,="8000 #1%
  \endgroup
}

@@

\begin{abstract}
\noindent
Direct reciprocity can explain how cooperation emerges in repeated social interactions. 
According to this literature, individuals should naturally learn to adopt conditionally cooperative strategies such as Tit-for-Tat. % if they expect to have further encounters with their interaction partner. 
Corresponding models have greatly facilitated our understanding of cooperation. 
Yet they often make strong assumptions on how individuals remember and process information. 
For example, when strategies are updated through social learning, it is commonly assumed that individuals update their strategies based on their average payoffs.  
This would require them to compute (or remember) their payoffs against all other population members.
Instead, herein, we introduce a theoretical framework to study the evolution of reciprocity when individuals learn based on their most recent experiences.
Even in the most extreme case that they only take into account their very last interaction, we find that cooperation can still evolve. 
However, such individuals adopt less generous strategies, and they tend to cooperate less often than in the classical framework with expected payoffs. 
Interestingly, once individuals remember the payoffs of two or three recent interactions, evolving cooperation rates quickly approach the classical limit. 
These findings contribute to a literature that explores which kind of cognitive capabilities are required for reciprocal cooperation. 
While our results suggest that memory facilitates the evolution of reciprocity, it only takes a rather modest amount of payoff memory for cooperation to emerge. 
\end{abstract}

@@

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\cite{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stankova:JAMA:2019}, and of cooperation~\citep{axelrod1981evolution}. 
Respective models translate strategic interactions into games~\cite{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies they can choose, and what fitness consequences (or payoffs) their strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
Models of biological evolution posit that individuals with a high fitness tend to produce more offspring; models of cultural evolution instead assume that such individuals are imitated more often. 
While biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies that are available, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may often depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, and on how easy payoff comparisons are.

@@

These difficulties -- to learn strategies by social imitation -- are particularly pronounced in models of direct reciprocity. 
This literature follows Trivers' insight that individuals have more of an incentive to cooperate in social dilemmas when they interact repeatedly~\citep{trivers1971evolution}. 
In repeated interactions, individuals can condition their future behavior on their past experiences with their interaction partner. 
They may use strategies such as Tit-for-Tat~\citep{rapoport:book:1965} or Generous Tit-for-Tat~\citep{molander:jcr:1985,Nowak1992tit} to preferentially cooperate with those partners who cooperated in the past. 
Such conditional strategies approximate human behavior fairly well~\citep{fischbacher:EconL:2001,Rand:TCS:2013,DalBo:AER:2019,Rossetti:ETH:2023} and they have also been documented in several other species~\citep{Carter:PRSB:2013,Schweinfurth:AnBehav:2019,Voelkl:PNAS:2015} -- although direct reciprocity is generally more difficult to demonstrate in animals~\citep{CluttonBrock:Nature:2009,Silk:CurrentBiology:2013,taborsky:CurrentBiology:2013}.
However, at the outset, it is not clear how easy it is to {\it learn} such strategies by social imitation. 
As one obstacle, individuals usually only observe other population members' actions (i.e., whether they cooperate or defect). 
These observations alone may not suffice to infer the underlying strategies (i.e., the contingent rules that determine whether to cooperate or defect in a given situation). 
As another obstacle, even if others' strategies are observable, individuals might find it difficult to identify which ones have the highest payoff. 
After all, the payoff of a strategy of direct reciprocity is not determined by the outcome of any single round.
Rather it is determined by how well this strategy fares over an entire sequence of rounds, against many different population members. 
In practice, such information might be both difficult to obtain and difficult to process.

@@

Most models of direct reciprocity abstract from these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}. 
Instead they assume individuals can easily copy the strategies of others. 
Similarly, they assume that updating decisions are based on the strategies' average (or expected) payoffs, taking into account all rounds and interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.][]{Hauert1997,van-veelen:PNAS:2012,Stewart2016,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning only based on payoff information from a limited number of rounds.

@@

To explore that question, we first consider two extreme scenarios. 
The first scenario corresponds to the usual modeling approach. 
Here, individuals update their strategies for a repeated prisoner's dilemma based on pairwise comparisons~\cite{traulsen2007pairwise}, based on their strategies' expected payoff. 
We contrast this model with an alternative scenario where individuals update their strategies based on the very
last (one-shot) payoff they obtained. 
We observe that individuals with limited payoff memory tend to adopt less generous strategies, yet moderate levels of cooperation can still evolve. 
Moreover, as we increase the individuals' payoff memory, to include the last two or three one-shot payoffs, cooperation rates quickly approach the rates observed in the classical baseline case. 
Overall, these findings suggest that while memory is important, already minimal payoff information may suffice for the evolution of direct reciprocity based on social learning. 
They also suggest that the classical model of reciprocity (based on expected payoffs) can be interpreted as a useful approximation to more realistic models that provide a more realistic depiction of cognitive constraints.

\section{Model Setup}\label{section:model}

A pairwise comparison process~\cite{traulsen2007pairwise} starts with assigning all
individuals of the population the same strategy. Thenceforth each elementary time step of
the process consists of the mutation phase, the game phase and the update phase.
In the game phase individuals are matched in pairs and that they participate in
a repeated 2 person donation game; a special case of the prisoner's dilemma. In
the donation game there are two actions: cooperation (\(C\)) and defection
(\(D\)). By cooperating a player provides a benefit \(b\) to the other player at
their cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn
are given by,

\begin{equation}
    \begin{blockarray}{ccc}
        & \text{cooperate} & \text{defect} \\
        \begin{block}{c(cc)}
            \text{cooperate} & b - c & -c \\
            \text{defect} & b & 0 \\
        \end{block}
    \end{blockarray}.
\end{equation}

Let \(\mathbf{u} = (b-c, -c, b, 0)\) be the round payoffs in a vector format, and let
\(\mathcal{U} = \{r, s, t, p\}\) denote the set of feasible payoffs, where \(r\)
denotes the payoff of mutual cooperation, \(s\) the sucker's payoff, \(t\) the
temptation to defect payoff, and \(p\) the punishment payoff.

In repeated games there are infinite many strategies, however, similar to the
literature we will assume that individuals use reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently. The play between a
pair of reactive strategies $s_1 = (y_1, p_1, q_1)$ and $s_2 = (y_2, p_2, q_2)$
can be model as a Markov process with the transition matrix \(M\)~\cite{nowak:APC:1989},

In the update stage two individuals are randomly selected. From the two
individuals, one  serves as the `learner' and the other as the `role model'. The
learner adopts the role model's strategy with a probability \(\rho\) given by,

\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{RM}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_{RM}\!-\! \pi_{L})}}.
\end{equation}

\(\pi_{L}\) and \(\pi_{RM}\) are the updating payoffs/fitness of the learner and the
role model respectively. The updating payoffs are a measure of how successful
individuals are in the current standing of the population. The parameter
\(\beta\) is known as the selection strength, namely, it shows how important the
payoff difference is when the learner is considering adopting the strategy of
the role model.

For the results presented here we assume that mutations are rare (\(\mu
\rightarrow 0\)). In fact, so rare that only two different strategies can be
present in the population at any given time. However, in the Supplementary Information
Section 9 we show that the main result holds for \(\mu \neq 0\).
The case of low mutation is vastly adopted because it allows us to explicitly
calculate the fixation probability of a newly introduced mutant. More specifically,
at each step
one individual adopts a mutant strategy randomly selected from the set of
feasible strategies. The fixation probability \(\phi_{M}\) of the mutant
strategy can be calculated explicitly as,

\section{Simulation Results}

To assess the impact of updating payoffs, we simulate the evolutionary process,
recording the strategies adopted by players over time based on perfect and
limited payoff memory. We performed two separate runs for each approach varying
the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (note that we omit the opening
move \(y\), as the discount factor \(\delta\) is relatively high). The figure
suggests that when updating is based on perfect payoff memory players tend to be
more generous and more cooperative.

@@

>>>>

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress,comma,round]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

\def\alld{\texttt{ALLD}}
\def\tft{\texttt{TFT}}
\def\gtft{\texttt{GTFT}}
\def\esm{electronic supplementary material}

@@

@@

\begin{abstract}
\noindent
Direct reciprocity can explain how cooperation emerges in repeated social interactions. 
According to this literature, individuals should naturally learn to adopt conditionally cooperative strategies, such as Tit-for-Tat. % if they expect to have further encounters with their interaction partner. 
Corresponding models have greatly facilitated our understanding of cooperation, yet they often make strong assumptions on how individuals remember and process information. 
For example, when strategies are updated through social learning, it is commonly assumed that individuals compare their respective average payoffs.  
This would require them to compute (or remember) their payoffs against all other population members.
Instead, herein, we introduce a theoretical framework to study the evolution of reciprocity when individuals learn based on their most recent experiences.
Even in the most extreme case that they only take into account their very last interaction, we find that cooperation can still evolve. 
However, such individuals adopt less generous strategies, and they tend to cooperate less often than in the classical setup with expected payoffs. 
Interestingly, once individuals remember the payoffs of two or three recent interactions, evolving cooperation rates quickly approach the classical limit. 
These findings contribute to a literature that explores which kind of cognitive capabilities are required for reciprocal cooperation. 
While our results suggest that some rudimentary form of payoff memory is necessary, it already suffices to remember a few interactions.\end{abstract}

@@

Evolutionary game theory describes the dynamics of populations when an individual's fitness depends on the traits or strategies of other population members~\cite{hofbauer1998evolutionary, nowak:Nature:2004, hauert2005game}.  
This theory can be used to describe the dynamics of animal conflict~\citep{maynard-smith:Nature:1973}, cancer cells~\citep{Stein:PTRS:2023}, and of cooperation~\citep{nowak:Science:2006}. 
Respective models translate strategic interactions into games~\cite{smith1982evolution}. 
These games specify how individuals (players) interact, which strategies they can choose, and what fitness consequences (or payoffs) their strategies have. 
In addition, these models also specify the mode by which successful strategies spread over time. 
Models of biological evolution posit that individuals with a high fitness tend to produce more offspring; models of cultural evolution instead assume that such individuals are imitated more often. 
Although biological and cultural evolution are sometimes treated as equivalent, there can be important differences~\citep{Wu2015,Smolla:PTRS:2021}. 
For example, models of biological evolution do not require individuals to have any particular cognitive abilities.
Here, it is the evolutionary process itself that biases the population towards strategies with higher fitness. 
In contrast, in models of cultural evolution, individuals need to be aware of the different strategies present in the population, and they need to identify those strategies with a higher payoff. 
As a consequence, evolutionary outcomes may depend on how easily different behaviors can be learned~\citep{Chatterjee:JTB:2012}, and on how easy payoff comparisons are.

@@

These difficulties -- to learn strategies by social imitation -- are particularly pronounced in models of direct reciprocity. 
This literature follows Trivers' insight that individuals have more of an incentive to cooperate in social dilemmas when they interact repeatedly~\citep{trivers1971evolution}. 
In repeated interactions, individuals can condition their future behavior on their past experiences with their interaction partner. 
They may use strategies such as Tit-for-Tat~\citep{rapoport:book:1965,axelrod1981evolution} or Generous Tit-for-Tat~\citep{molander:jcr:1985,Nowak1992tit} to preferentially cooperate with those partners who cooperated in the past. 
Such conditional strategies approximate human behavior fairly well~\citep{fischbacher:EconL:2001,Rand:TCS:2013,DalBo:AER:2019,Rossetti:ETH:2023} and they have also been documented in several other species~\citep{Carter:PRSB:2013,Schweinfurth:AnBehav:2019,Voelkl:PNAS:2015} -- although direct reciprocity is generally more difficult to demonstrate in animals~\citep{CluttonBrock:Nature:2009,Silk:CurrentBiology:2013,taborsky:CurrentBiology:2013}.
However, at the outset, it is not clear how easy it is to {\it learn} such strategies by social imitation. 
As one obstacle, individuals usually only observe other population members' actions (i.e., whether they cooperate or defect). 
These observations alone may not suffice to infer the underlying strategies (i.e., the contingent rules that determine whether to cooperate or defect in a given situation). 
As another obstacle, even if others' strategies are observable, individuals might find it difficult to identify which ones have the highest payoff. 
After all, the payoff of a strategy of direct reciprocity is not determined by the outcome of any single round.
Rather it is determined by how well this strategy fares over an entire sequence of rounds, against many different population members. 
In practice, such information might be both difficult to obtain and difficult to process.

@@

Most models of direct reciprocity abstract from these difficulties~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}. 
Instead they assume individuals can easily copy the strategies of others. 
Similarly, they assume that updating decisions are based on the strategies' average (or expected) payoffs, taking into account all rounds and interactions. 
These assumptions create a curious inconsistency in how models represent an individual's cognitive abilities. 
On the one hand, when playing the game, individuals are often assumed to have restricted memory. 
Respective studies typically assume that individuals make their decisions each round based on the outcome of the last round only~\citep[with only a few exceptions, see Refs.][]{Hauert1997,van-veelen:PNAS:2012,Stewart2016,Li:NatCS:2022,Murase:PLoSCompBio:2023a}. 
Yet when learning new strategies, individuals are assumed to remember (or compute) each others' precise average payoff across many rounds and many interaction partners. 
Herein, we wish to explore whether this latter assumption is actually necessary for the evolution of reciprocity through social imitation. 
We ask whether individuals can learn to adopt reciprocal strategies even when learning is based on payoff information from a limited number of rounds.

@@

To explore that question, we first consider two extreme scenarios. 
The first scenario corresponds to the usual modeling approach. 
Here, individuals update their strategies for a repeated prisoner's dilemma based on pairwise comparisons~\cite{traulsen2007pairwise}, taking into account their expected payoffs. 
We contrast this model with an alternative scenario where individuals update their strategies based on the very
last (one-shot) payoff they obtained. 
We find that individuals with limited payoff memory tend to adopt less generous strategies, yet moderate levels of cooperation can still evolve. 
Moreover, as we increase the individuals' payoff memory to include the last two or three one-shot payoffs, cooperation rates quickly approach the rates observed in the classical baseline case. 
Overall, these findings suggest that while memory is important, already minimal payoff information may suffice for the evolution of direct reciprocity based on social learning. 
They also suggest that the classical model of reciprocity (based on expected payoffs) can often be interpreted as a useful approximation to more realistic models that include cognitive constraints.

\section{Model and Methods}\label{section:model}

%% MODEL: Basic intro

To explore the impact of limited payoff memory, we adapt existing models of the evolution of direct reciprocity through social learning. 
These models consider the dynamics on two different time scales. 
The short time scale describes the game dynamics. 
Here, individuals with fixed strategies are randomly matched to interact with each other in a series of repeated social dilemmas. 
The long time scale describes the evolutionary dynamics. 
Here, individuals can update their strategies based on the payoffs they yield. 
In the following, we describe the basic setup of our model; all details and derivations are described in the \esm.\\

%% MODEL: (Repeated) donation game

\noindent
{\bf Description of the game dynamics.} For our model, we consider a well-mixed population consisting of $N$~players.
Players are randomly matched in pairs to participate in a repeated donation game~\citep{sigmund2010calculus}.
Each round, players can either cooperate (\(C\)) or defect (\(D\)). 
By cooperating, a player provides a benefit \(b\) to the other player at their cost \(c\), with \(0 \!<\! c \!<\! b\). 
Thus, the players' payoffs in a single round are given by the matrix

\begin{equation}
    \begin{blockarray}{ccc}
        & \text{cooperate} & \text{defect} \\
        \begin{block}{c(cc)}
            \text{cooperate} & b - c & -c \\
            \text{defect} & b & 0 \\
        \end{block}
    \end{blockarray}.
\end{equation}
In particular, payoffs take the form of a prisoner's dilemma.
Mutual cooperation yields a better payoff than mutual defection ($b\!-\!c\!>\!0$), but each player individually prefers to defect independent of the co-player's action ($b\!>\!b\!-\!c$ and $0\!>\!-c$). 
To incorporate that individuals interact repeatedly, we assume that after each round, there is a constant continuation probability $\delta$ of interacting for another round. 
For $\delta\!=\!0$, we recover the case of a conventional (one-shot) prisoner's dilemma. 
Here, mutual defection is the only equilibrium. 
As $\delta$ increases, the game turns into a repeated game. As a result, additional equilibria emerge, with some of them allowing for full cooperation~\citep{friedman:RES:1971,Akin:chapter:2016,hilbe:GEB:2015,stewart:pnas:2014}.

%% MODEL: Reactive strategies

In a one-shot donation game, players can only choose among two strategies (either cooperate or defect).
In the repeated game, strategies can become arbitrarily complex. 
Here, strategies are contingent rules, telling players what to do depending on the outcome of all their previous interactions. 
For simplicity, in the following we assume individuals use {\it reactive strategies}~\citep{Nowak1992tit}. 
A reactive strategy only depends on the other player's action in the previous round. 
As a consequence, reactive strategies can be written as a three-dimensional tuple \(\mathbf{s}=(y, p,q)\). 
The first entry \(y\) is the probability that the player opens with cooperation in the first round. 
The two other entries are the probabilities that the player cooperates in all subsequent rounds, depending on whether the co-player cooperated~($p$) or defected~($q$) in the previous round. 
The set of reactive strategies is simple enough to allow for an explicit mathematical treatment~\citep{hofbauer1998evolutionary}. 
Yet it is rich enough to capture several important strategies of repeated games. 
For example, it contains \alld{} $=\!(0,0,0)$, the strategy that always defects. 
Similarly, it contains Tit-for-Tat, \tft{} $=\!(1,1,0)$, the strategy that copies the co-player's previous action (and that cooperates in the first round).

In the short run, the players' strategies are taken to be fixed.
Players use their strategies to decide whether to cooperate with any given interaction partner. 
In the long run, however, the players' strategies may change depending on the payoffs they yield, as we describe in the following.\\

 %% MODEL: Pairwise comparison process

\noindent
{\bf Description of the evolutionary dynamics.}
Herein, we assume population members update their strategies based on social learning. 
To model these strategy updates, we consider a pairwise comparison process~\citep{traulsen2007pairwise}. 
This process assumes that at regular time intervals, one population member is randomly selected, and given the chance to update its strategy.
We refer to this player as the `learner'. 
With probability $\mu$ (reflecting a mutation rate), the learner simply adopts a random strategy (with all reactive strategies having the same probability to be chosen). 
With the converse probability $1\!-\!\mu$, the learner randomly picks a `role model' from the population. 
The learner then compares its own payoff $\pi_L$ from the repeated game stage to the payoff $\pi_R$ of the role model. 
The learner adopts the role model's strategy with a probability \(\rho\) described by a Fermi function~\citep{blume:GEB:1995,szabo:PRE:1998}, 
\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{R}) = \frac{1}{1\!+\! e^{- \!\beta (\pi_{R}- \pi_{L})}}.
\end{equation}
Here, the selection strength $\beta\!\ge\!0$ is a parameter that indicates how sensitive players are to payoff differences. 
For $\beta\!=\!0$, any payoff differences are irrelevant, and the learner simply adopts the role model's strategy with probability one half. As the selection strength~$\beta$ increases, players are increasingly biased to update only if the role model has the higher payoff. 
In either case, we assume that the role model's strategy can be identified.
That is, if the learner decides to imitate the role model's strategy, the learner copies it exactly, as assumed in most previous work on direct reciprocity.

%% MODEL: Perfect vs limited payoff memory

We deviate from previous models in how we interpret the payoffs $\pi_L$ and $\pi_R$, which form the basis of the pairwise comparisons in Eq.~\eqref{Eq:rho}. 
In previous work, these payoffs are taken to be the respective players' expected payoffs. 
We interpret this setup as a model with perfect payoff memory. 
In this setup, the payoffs  $\pi_L$ and $\pi_R$ represent an average over all possible repeated games the two individuals could have played with all population members. 
The use of expected payoffs is mathematically convenient, because explicit formulas for the players' expected payoffs are available~\citep{hofbauer1998evolutionary}.
Herein, we compare this model to a model with limited payoff memory. 
In that model, the players' payoffs $\pi_L$ and $\pi_R$ are taken to be the payoffs that each player received in the last round of the last repeated game they interacted in (see \FigEvoProc\textbf{a}). 
This assumption could reflect, for example, a strong recency bias in how individuals evaluate their payoffs.  
In addition to this extreme case of limited payoff memory, later on we also explore cases in which players take into account the outcome of two, three, or four recent rounds.

Both in the case of perfect and limited memory, we iterate this elementary pairwise comparison process for many time steps. 
This gives rise to a stochastic process that describes which strategies players adopt over time. 
We explore the dynamics of this process with computer simulations.
For the results presented in the following, we assume that mutations are rare (\(\mu\!\rightarrow\! 0\)). 
This assumption is fairly common in evolutionary game theory, because it makes computations more efficient~\citep{fudenberg:JET:2006,wu:JMB:2012,mcavoy:jet:2015}, and because the results can be interpreted more easily.  
However, in Section~9 of the \esm{} we show that our main result holds continue to hold for strictly positive mutation rates.

\section{Results}

\noindent
{\bf Stability of cooperative populations.}
To get some intuition for the differences between perfect and limited payoff memory, we first analyze when cooperation is stable in either scenario.
To this end, we consider a resident population in which all players but one adopt a strategy of Generous Tit-for-Tat, \gtft{} $=\!(1,1,q)$. 
The remaining mutant player adopts \alld. 
We say {\it cooperation is stochastically stable} if the single mutant is more likely to imitate the residents than vice versa. 
For simplicity, we consider a large population~($N\!\rightarrow\!\infty$) and strong selection~($\beta\!\rightarrow\!\infty$).
More general results are derived in the \esm.

In the case of perfect payoff memory, it is straightforward to compute each player's expected payoff. 
Because the population mostly consists of residents, and because residents mutually cooperate with each other, their payoff is $\pi_\gtft = b\!-\!c$. 
The mutant only interacts with residents, and hence receives a benefit in the first round, and in every subsequent round with probability $q$. 
As a result, the mutant's expected payoff is $\pi_\alld \!=\! (1\!-\!\delta\!+\!\delta q)b$. 
For perfect payoff memory, the requirement for cooperation to be stochastically stable reduces to the condition $\pi_\gtft > \pi_\alld$. 
This yields
\begin{equation} \label{Eq:PerfectMemory}
q < 1\!-\!\frac{c}{\delta  b}.
\end{equation}
In particular, we recover that $q\!=\!1-c/(\delta b)$ is the maximum generosity that cooperators should have~\citep{molander:jcr:1985,Nowak1992tit}. 
Because $q$ needs to be non-negative, we also conclude that cooperation can only be stable if $\delta \!>\! c/b$.
Again, this condition for the feasibility of direct reciprocity is the condition found in the literature~\citep{nowak:Science:2006}.

The logic of the case with limited payoff memory is somewhat different. 
Here we need to compute how likely each player obtains one of the four possible payoffs $\{b\!-\!c, -c, b, 0\}$ in their last interaction. 
Because residents almost always interact with other residents, their last one-shot payoff is $\pi_\gtft = b\!-\!c$ with a probability approaching one. 
For the defecting mutant, there are two possibilities. 
({\it i})~ If the mutant's co-player happens to cooperate in the last round, the mutant receives a payoff of $\pi_\alld\!=\!b$.
This occurs with probability $1\!-\!\delta\!+\!\delta q$. 
({\it ii})~If the co-player defects in the last round, the mutant receives $\pi_\alld=\!0$. 
This case occurs with probability $\delta(1\!-\!q)$.
Because $b\!-\!c\!<\!b$, we conclude that  in the first case, residents tend to imitate the mutant. 
Because $b\!-\!c\!>\!0$, in the second case mutants tend to imitate the resident. 
Cooperation is thus stochastically stable if the first case is less likely than the second. 
This yields the condition
\begin{equation}
q < 1\!-\!\frac{1}{2 \delta}.
\end{equation}
Interestingly, this condition no longer depends on the exact payoff values $c$ and $b$ (this happens because selection is strong, such that only the payoff ordering $b\!>\!c\!>\!0$ matters). In particular, because $q$ is non-negative, this condition can only be satisfied if $\delta\!>\!1/2$.

By comparing the two cases, we find that payoff memory affects which cooperative strategies are viable. 
With perfect memory, the maximum generosity $q$ needs to satisfy Eq.~\eqref{Eq:PerfectMemory}.
In particular, this generosity can become arbitrarily large, if only the game's benefit-to-cost ratio~$b/c$ and the continuation probability~$\delta$ are sufficiently large. 
In contrast, with limited payoff memory, the maximum generosity is bounded by one half, and it is independent of the benefit-to-cost ratio. 
To explore whether we indeed observe these patterns in evolving populations, we turn to simulations.\\

\noindent
{\bf Evolutionary dynamics with perfect and limited payoff memory.}
To assess the impact of updating payoffs, we simulate the evolutionary process,
recording the strategies adopted by players over time based on perfect and
limited payoff memory. We performed two separate runs for each approach varying
the value of benefit \(b\).
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$ (note that we omit the opening
move \(y\), as the discount factor \(\delta\) is relatively high). The figure
suggests that when updating is based on perfect payoff memory players tend to be
more generous and more cooperative.

@@

%% SOME ADDITIONAL TEXT BITS %%

% repeated game markov chain
given that the opponent cooperated and defected equivalently. The play between a
pair of reactive strategies $s_1 = (y_1, p_1, q_1)$ and $s_2 = (y_2, p_2, q_2)$
can be model as a Markov process with the transition matrix \(M\)~\cite{nowak:APC:1989},

% pairwise comparison fixation probability

In fact, so rare that only two different strategies can be
present in the population at any given time. However, in the Supplementary Information
Section 9 we show that the main result holds for \(\mu \neq 0\).
The case of low mutation is vastly adopted because it allows us to explicitly
calculate the fixation probability of a newly introduced mutant. More specifically,
at each step
one individual adopts a mutant strategy randomly selected from the set of
feasible strategies. The fixation probability \(\phi_{M}\) of the mutant
strategy can be calculated explicitly as,

% Explicit formulas for the transition probabilities for perfect and limited memory
